Random Matrix Theory Tools for the Predictive Analysis of Functional Magnetic Resonance Imaging Examinations

### 2023-04-15 paper/paper/SPIE/response_letter.tex

%%%%%% 16. Line 311: I see what they mean by "voxel time series" because of lines
%%%%%% 317-318, so it might be clearer to move those lines to directly after line 311,
%%%%%% so it is clear without having to read further.

17. The authors do a nice job of presenting the relevant results, but the
conclusions were rather general, being along the lines of "investigate the
sensitivity of choices made in the processing and analysis". That's an
important result, given previous work and shows that RMT might not "work" with
different data sets and processing workflows. The question is: what's the
takeaway: that the method is not one in which you can have a default set of
parameters that will generally give good results (like say, the recon-all
pipeline for FreeSurfer)? If RMT so sensitive to preprocessing and choice of
parameters, is this a useful method to be applied widely on fMRI data? It would
be good for this to be discussed and perhaps some recommendations given.
}

\noindent
\today, \\
Dear Dr. Gimi, \\

Thank you for your careful consideration of our manuscript
\emph{Random Matrix Theory Tools for the Predictive Analysis of Functional
Magnetic Resonance Imaging Examinations} submitted for publication in SPIE
Journal of Medical Imaging. We especially thank our anonymous reviewer,
who and offered especially constructive and specific comments on this manuscript.

The updated manuscript indicates changes with coloration, as suggested by
SPIE:JMI. Minor wording or other uncontroversial changes are identified in
\textcolor{orange}{orange} in the updated .pdf manuscript (or with
\verb|\textcolor{orange}{text}| in the \texttt{.tex} file). More significant
changes are identified in \textcolor{cyan}{cyan} (\verb|\textcolor{cyan}{text}|).

\begin{itemize}
\item Reviewer comments 2, 4, 5, 6, 10, 11, 13, 14, 15, 16 noted easily-correctable
and uncontroversial issues relating to phrasing, clarity or ordering, and we
have made changes in line with those comments on lines 55, 115, 136, 153,
196, 224, 247, 381-387, 413 and 309, and 329-331, respectively.

\item Comment 3 interprets our statement as making a broad claim, but the statement
was in fact intended to qualify or limit the scope of applicability of RMT.
This has been made clearer (lines 69-71). Likewise, we don't think there is
any disagreement between our intended meaning and the sentiment in comment 7,
and have revised line 159 to make it clear that our doubts are confined to
\textit{past} studies.

\item Reviewer comments 8 and 9 noted some issues with vagueness involving
descriptions of the fMRI scans and runs in the text and tables. We hope these
descriptions are now unambiguous (lines 169-174 and Table 1 caption).

\item In response to comment 12, while we did, and still do take as given that our
reviewer and a reader of SPIE:JMI knows what features and feature extraction
are, we can see how at this point in the paper, it may not be clear
\textit{why} we are extracting features. We have added a short paragraph (lines
234-241) that we hope resolves this, and makes it clear we are comparing the
predictive utility of different features / feature extraction methods (and that
``feature'' has the usual meaning in this context).
\end{itemize}

In response to our reviewer's first comment, we have expanded the introduction
to include both the definition of a random matrix, and some concrete examples of
random matrix ensembles or classes, and also a brief explanation of how RMT
theory is used in empirical contexts. We hope that this covers the basics both
in a way that presumes neither too little nor too much from readers (a very
difficult balance to strike with something like RMT, unfortunately).

Our reviewer's final comment suggested (by our reading) to give our conclusion
a bit more punch. The point is well-taken, and we have reworked our conclusion
to more strongly imply a need for caution and/or carefulness in the
presentation and interpretation of RMT-fMRI-related findings.

Please address all correspondence concerning this manuscript to \href{jlevman@stfx.ca}{jlevman@stfx.ca}.

### 2023-04-15 paper/paper/SPIE/article.tex
mostly finalize review response

\color{cyan}

At its most rudimentary, the basic insight of Random Matrix Theory (RMT) is
that matrices can be described by their eigenvalues, and that
large matrices (or large collections of matrices) can be
well-described by the statistical behaviour of those eigenvalues. Such a
description is obtained by assuming that the large matrix (or collection
of matrices) is sampled from a \textit{class} or \textit{ensemble} of random
matrices.

Simplifying somewhat, a random matrix is a matrix that has entries that can be
well-described as being drawn from a particular independent and identically
distributed (\textit{iid}) distribution, or, equivalently, is a matrix where
each entry is itself a random variable. A \textit{class} or \textit{ensemble}
of random matrices is generally defined by the shape of the \textit{iid}
distribution involved. For example, the Gaussian Orthogonal Ensemble (GOE) is a
\textit{class} or \textit{ensemble} of random matrices that comprises
orthogonal matrices with each entry being sampled from a standard Gaussian, and
the Marchenko-Pastur distribution describes the limiting behaviour of the
singular values of \textit{iid} rectangular random matrices where the
\textit{iid} distribution is arbitrary\cite{mehtaRandomMatrices2004}. Of
course, given a single particular matrix, one can only estimate a
\textiit{probability} that it is a member of a certain class, since certain RMT
classes can yield arbitrary matrices (albeit with extremely low probability).

and eigenvalue distributions of GOE random matrices have a
number of well-understood properties\cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}. Even more generally,

Specifically, RMT finds that often, in the infinite limit, that the expected
distribution of the eigenvalues of a number of classes of random matrices can
be described quite precisely.\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}. Although these
assumptions are quite specific and restrictive, nevertheless, RMT-predicted
eigenvalue distributions have appeared in diverse natural phenomena.

@@

\textcolor{cyan}{Thus, while} first developed to describe the fluctuation of nuclei
energy levels in quantum physics \cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}, RMT has been shown to have extremely
broad potential. In small-scale physical systems, RMT universalities have been
observed in quantum chaotic systems, complex nuclei, atoms, molecules and
disordered mesoscopic systems \cite{guhrRandommatrixTheoriesQuantum1998a,
mehtaRandomMatrices2004, brodyRandommatrixPhysicsSpectrum1981,
beenakkerRandommatrixTheoryQuantum1997,
bohigasHigherOrderCorrelationsSpectra1985, wintgenLevelStatisticsQuantized1988,
pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has
been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002}, social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids\cite{sastrySpectralStatisticsInstantaneous2001,
matharooSpectralStatisticsQuenched2009} and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,
matharooVibrationalSpectraAmorphous2005,
matharooUniversalityVibrationalSpectra2008}. Within biological systems, RMT has
also been used to successfully model aspects of  amino acid functional
relationships \cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in
epileptic seizures \cite{osorioPhasesynchronizationRandommatrixBased2011}, and
in protein-protein interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep neural
networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

\textcolor{orange}{When or if RMT has} real-world explanatory potential,
\textcolor{orange}{this will most likely be} when dealing with a complex system
of many (hundreds or more\cite{mehtaRandomMatrices2004}) interacting
components, \textcolor{orange}{rather than a system that is too small for
statistical regularities ro reliably surface}. If such a system has a matrix
representation, and the eigenvalues of this representation have a distribution
similar to one predicted by RMT, it suggests the system is either highly random
or chaotic. By contrast, if the observed spectra deviate significantly from
RMT-predicted spectra, this
suggests otherwise. A number of studies have used RMT to make such
interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

The \textcolor{orange}{computing} time and \textcolor{orange}{memory
requirements} for calculating the full \(N \times N\) \(\mathbf{M}\) matrix of
Pearson correlation coefficients (and subsequent eigenvalues) are too large to
be tractable. However, since transposition does not change the eigenvalues, we
can use the transpose to efficiently compute these eigenvalues (see Appendix
\ref{sec:transpose}). These raw eigenvalues (feature \code{eigs} in Table
\ref{tab:features}), most directly test the idea that eigenvalues alone are
useful functional connectivity features.

@@

For combined features, slicing variants are also computed, with slicing
performed first on each feature to be combined. E.g. if we have features \(f_1,
\dots,  f_n\) with dimensions \(p_1, \dots, p_n\), then the sliced max 25\%
feature for each component feature, \(\hat{f}_i\) is the last \(p_i / 4\)
elements of \(f_i\), and the combined sliced feature is  \([\hat{f}_1;\;
\dots;\; \hat{f}_n]\).
% \footnote{Exact implementation can be found in the
%\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated\_predict.py\#L182-L213}{source code}}.

@@

\textcolor{cyan}{Because the computation of these statistics is
non-trivial, and to make the unfolding procedure more transparent and
reproducible, we developed and release a separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances. The library also makes available
various other functions useful for empirical RMT analyses, such as unfolding
and trimming functions, and plotting facilities for classic RMT ensembles.}

@@

Eigenvalue-based features inspired by RMT, and extracted from fMRI functional
connectivity matrices were found to have predictive utility across a wide
variety of datasets and classification tasks. However, the predictive utilities
were modest, and highly dependent on preprocessing steps and other fitting and
feature selection procedures. Future studies employing RMT should carefully
investigate the sensitivity of any findings to such analytic decisions before
drawing any conclusions. However, this study demonstrates that the inclusion of
RMT statistics and other eigenfeatures extracted from fMRI images could aid in
the predictive modeling of a wide variety of phenomena.

>>>>

Though first developed to describe the fluctuation of nuclei energy levels in
quantum physics \cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}, Random Matrix Theory (RMT) has been
shown to have extremely broad potential. In small-scale physical systems, RMT
universalities have been observed in quantum chaotic systems, complex nuclei,
atoms, molecules and disordered mesoscopic systems
\cite{guhrRandommatrixTheoriesQuantum1998a, mehtaRandomMatrices2004,
brodyRandommatrixPhysicsSpectrum1981, beenakkerRandommatrixTheoryQuantum1997,
bohigasHigherOrderCorrelationsSpectra1985, wintgenLevelStatisticsQuantized1988,
pandeySkewOrthogonalPolynomialsUniversality2001}, and at larger scales, RMT has
been applied to atmospheric physics
\cite{santhanamStatisticsAtmosphericCorrelations2001}, stock cross-correlations
\cite{plerouRandomMatrixApproach2002}, social networks
\cite{jalanUncoveringRandomnessSuccess2014}, random networks
\cite{bandyopadhyayUniversalityComplexNetworks2007}, network-formation in
liquids\cite{sastrySpectralStatisticsInstantaneous2001,
matharooSpectralStatisticsQuenched2009} and amorphous clusters
\cite{sarkarUniversalityVibrationalSpectra2004,
matharooVibrationalSpectraAmorphous2005,
matharooUniversalityVibrationalSpectra2008}. Within biological systems, RMT has
also been used to successfully model aspects of  amino acid functional
relationships \cite{bhadolaTargetingFunctionalMotifs2016}, synchrony in
epileptic seizures \cite{osorioPhasesynchronizationRandommatrixBased2011}, and
in protein-protein interactions both in different species
\cite{agrawalQuantifyingRandomnessProtein2014} and breast cancer
\cite{raiRandomnessPreservedPatterns2015}. RMT has also been used to guide
statistical decisions in principal components
analyses\cite{franklinParallelAnalysisMethod1995,
veraartDenoisingDiffusionMRI2016, ulfarssonDimensionEstimationNoisy2008}, and,
more recently, has provided insights into the behaviors of deep neural
networks\cite{martinImplicitSelfRegularizationDeep2021,
martinPredictingTrendsQuality2021}.

\color{cyan}
At its most rudimentary, RMT describes the expected behaviour of the
eigenvalues—also often called \textcolor{orange}{the \textit{spectra} or}
\textit{levels}—of a number of classes of random
matrices\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}.  A random matrix is
a matrix where each entry is a random variable. Typically, each random variable
is \textcolor{orange}{independent} and identically distributed (\textit{iid}), however, modern
extensions allow for some
dependence\cite{bouchaudFinancialApplicationsRandom2009}. A \textit{class} or
\textit{ensemble} of random matrices can be defined by the type of
\textit{iid} distribution involved. For example, the Gaussian Orthogonal
Ensemble (GOE\cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}) is a \textit{class} or \textit{ensemble}
of random matrices that comprises orthogonal matrices with each entry being
sampled from a standard Gaussian, and the Marchenko-Pastur distribution
describes the limiting behaviour of the singular values of \textit{iid}
rectangular random matrices where the \textit{iid} distribution is
arbitrary\cite{mehtaRandomMatrices2004}. RMT often finds that in the infinite
limit, the expected eigenvalue distribution of a number of classes of random
matrices can be described quite
precisely\cite{andersonIntroductionRandomMatrices2010,
akemannOxfordHandbookRandom2011, mehtaRandomMatrices2004}.

Of course, given a particular non-random matrix or instantiation of a random
matrix, one can, in general, only estimate the likelihood that it is a member
of a certain class or ensemble, since certain RMT ensembles (such as the GOE)
can yield arbitrary matrices with extremely low probability. Thus, RMT also
includes statistical measures ("spectral observables") and tools to help
compare empirical observations to theory\cite{mehtaRandomMatrices2004,
guhrRandommatrixTheoriesQuantum1998a}.

% At its most rudimentary, the basic insight of Random Matrix Theory (RMT) is
% that matrices can be described by their eigenvalues, and that
% large matrices (or large collections of matrices) can be
% well-described by the statistical behaviour of those eigenvalues.

@@

\textcolor{orange}{When or if RMT has} real-world explanatory potential,
\textcolor{orange}{this will most likely be} when dealing with a complex system
of many (hundreds or more\cite{mehtaRandomMatrices2004}) interacting
components, \textcolor{orange}{rather than a system that is too small for
statistical regularities ro reliably surface}. If such a system has a matrix
representation, and the eigenvalues \textcolor{orange}{and spectral
observables} of this representation have distributions similar to
\textcolor{orange}{those} predicted by RMT, it suggests the system is either
highly random or chaotic. By contrast, if the observed spectra deviate
significantly from RMT-predicted spectra, this suggests otherwise. A number of
studies have used RMT to make such interpretations and comparisons between
systems\cite{santhanamStatisticsAtmosphericCorrelations2001,
jalanUncoveringRandomnessSuccess2014,
bandyopadhyayUniversalityComplexNetworks2007,
agrawalQuantifyingRandomnessProtein2014, raiRandomnessPreservedPatterns2015,
sebaRandomMatrixAnalysis2003, wangSpectralPropertiesTemporal2015,
wangRandomMatrixTheory2016, matharooSpontaneousBackpainAlters2020}.

@@

The \textcolor{orange}{computing} time and \textcolor{orange}{memory
requirements} for calculating the full \(N \times N\) matrix \(\mathbf{M}\) of
Pearson correlation coefficients (and subsequent eigenvalues) are too large to
be tractable. However, since transposition does not change the eigenvalues, we
can use the transpose to efficiently compute these eigenvalues (see Appendix
\ref{sec:transpose}). These raw eigenvalues (feature \code{eigs} in Table
\ref{tab:features}), most directly test the idea that eigenvalues alone are
useful functional connectivity features.

@@

For combined features, slicing variants are also computed, with slicing
performed first on each feature to be combined. \textcolor{orange}{For example}
if we have features \(f_1, \dots,  f_n\) with dimensions \(p_1, \dots, p_n\),
then the sliced max 25\% feature for each component feature, \(\hat{f}_i\) is
the last \(p_i / 4\) elements of \(f_i\), and the combined sliced feature is
\([\hat{f}_1;\; \dots;\; \hat{f}_n]\).
% \footnote{Exact implementation can be found in the
%\href{https://github.com/DM-Berger/random-matrix-fmri/blob/7c9e4187f582dedee728cd7193b8894d928c2f00/code/rmt/updated\_predict.py\#L182-L213}{source
%code}}.

@@

\textcolor{orange}{Because the computation of these statistics is
non-trivial, and to make the unfolding procedure more transparent and
reproducible, we developed and release a separate, open-source Python library
\code{empyricalRMT}\cite{dm-bergerStfxecutablesEmpyricalRMTV12022}. The library
allows for efficient, parallel computation of the spectral rigidity and level
variance via Monte Carlo methods, and automatically ensures convergence of the
statistics to user-specified tolerances. The library also makes available
various other functions useful for empirical RMT analyses, such as unfolding
and trimming functions, and plotting facilities for classic RMT ensembles.}

@@

Eigenvalue-based features inspired by RMT, and extracted from fMRI functional
connectivity matrices were found to have predictive utility across a wide
variety of datasets and classification tasks. However, the predictive utilities
were modest, and highly dependent on preprocessing steps and other fitting and
feature selection procedures.

\textcolor{cyan}{Given the sensitivity to these decisions observed in this
study, and considering the lack of consensus regarding the preprocessing of
fMRI or other complex biological signal data, RMT should probably not be
currently considered a tool with "out-of-the-box" utility in these domains.
While RMT likely still has potential to yield insights in a variety of
contexts, our results suggest that, absent strong evidence otherwise, it is not
safe to assume that these insights will generalize broadly beyond the specific
preprocessing and analytic pipelines involved.}

\textcolor{cyan}{Further research might establish specific sets of analytic
choices that allow RMT to consistently extract useful information in a wide
variety of contexts. However, it is also possible that each unique context
might require a specific combination of choices. In this latter case, there
should be \textit{strong} theoretical justification for the use of
RMT, and for each of the various analytic choices involved in its use: the
variability of results seen in this study suggests that there may often be a
set of analytic choices that provide favourable results, and which can be
weakly justified by plausible, but ultimately \textit{post hoc} justifications.}
\textcolor{cyan}{Provided that these precautions are followed, and that} future
studies employing RMT \textcolor{orange}{also} carefully investigate the
sensitivity of any findings to such analytic decisions, \textcolor{cyan}{then
there is likely considerable untapped potential for RMT in the analysis of fMRI.}
